---
title: "Comparing means from multiple groups of data"
output:
  html_document:
    df_print: paged
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```
_Acknowledgements: I would really like to thank Alboukadel Kassambara for his excellent rstatix package and for the great tutorial https://www.datanovia.com/en/courses/comparing-multiple-means-in-r/ _

# Part One: Parametric tests without repeated measures

Probably you are familiar with the t-test for comparing the mean of two groups. Here we discuss what to do when you want to compare the mean of multiple groups. Typically this will have two steps:  

1. an omnibus test e.g. a one-way ANOVA to see if you can reject the hypothesis that the groups are the same
2. a post-hoc test where you adjust for multiple comparisons and identify which groups are different.

#### Datasets used in this tutorial
```{r}
library(tidyverse)
set.seed(5)
datapoints <- 30
df <- tibble(
  temperature = rep(c(15,20,25,30),each = datapoints),
  activity = c(rnorm(datapoints, mean = 5, sd = 2), 
               rnorm(datapoints, mean = 5.5, sd = 2),
               rnorm(datapoints, mean = 11, sd = 2),
               rnorm(datapoints, mean = 9, sd = 2))
)
df$temperature <- as.factor(df$temperature)

set.seed(5)
dataset_treatment <- tibble(
  treatment = rep(c('control','compoundA','compoundB','compundC'),each = datapoints),
  activity = c(rnorm(datapoints, mean = 5, sd = 2), 
               rnorm(datapoints, mean = 5.5, sd = 2),
               rnorm(datapoints, mean = 11, sd = 2),
               rnorm(datapoints, mean = 9, sd = 2))
)
dataset_treatment$treatment <- as.factor(dataset_treatment$treatment)


```


## One-way ANOVA and Welch's ANOVA

### Summary checklist:

1. Repeated Measures? YES- use something else
2. Data approximately normally distributed?  
    - Check with Q-Q plot and/or Shapiro-Wilk test
    - If not: use a _non-parametric test_
3. Variance of the data in each group approximately equal?
    - If not: use _Welch’s ANOVA_ 
4. Now conduct the One-way ANOVA
5. Can you reject the hypothesis that all groups have equal means? If so carry on to a post-hoc test to find out which groups are different:
    - Do you want to compare all the groups with each other? Use the _Tukey_ test.
    - Do you want to compare multiple treatment groups with one control group? Use the _Dunnett_ test.

However if you decide to use Welch's ANOVA given unequal variance between groups:

4. Conduct Welch's ANOVA
5. Use appropriate post-hoc tests


### Assumptions of one-way ANOVA:

1. independent observations - this is not for "repeated measures" 
2. data for each group should be approximately normally distributed
3. variance of the data for each group should be equal.


#### Test for normality:

1. Q-Q plot should be roughly linear
```{r}
qqnorm(df$activity)
```

2. Shapiro-Wilk test of normality for all the data grouped together (however for very large datasets even small deviations from normality may be significant in this test)
```{r}
library(rstatix)
model  <- lm(activity ~ temperature, data = df)
shapiro_test(residuals(model))
```
As the P-value is > 0.05 cannot reject the hypothesis that the data is normally distributed.

3. If you have a decent number of points for each group, you can test each group separately for normality:
```{r}
df %>%
  group_by(temperature) %>%
  shapiro_test(activity)
```
or plot the Q-Q plot for each group
```{r}
library(ggpubr)
ggqqplot(df, "activity", facet.by="temperature")
```

#### Check that the variance is approximately the same for each group

Inspect with a boxplot:
```{r}
ggplot(df, aes(x=temperature, y=activity)) +
  geom_boxplot()
```


Residuals plot: 
```{r}
plot(model,1)
```

Or Levene's test (more robust to departures from normality than the Bartlett’s test). 
Read more about this here: https://uc-r.github.io/assumptions_homogeneity#levene

```{r}
df %>% levene_test(activity ~ temperature)
```
Can't reject the hypothesis that the variance of each group is the same.

## Conducting the One-Way ANOVA
remember is it dependent variable ~ independent variable
```{r}
anova.results <- df %>% anova_test(activity ~ temperature)
anova.results
```
P < 0.05 therefore we reject the null hypothesis that all groups are the same.
How to report this in a paper? 
```{r}
paste0('[F(', anova.results$DFn,',',anova.results$DFd,') = ',
       anova.results$F, ', p = ',anova.results$p,']')
```
So write something like: A one-way ANOVA found that there was a significant effect of temperature on activity then add from above [F(n,n=xxx, p = xxx)] 

## Post-hoc test after one-way ANOVA
You need to use a post-hoc test that takes account of the fact that you are making multiple comparisons, you can't just do several t-tests.

### For the situation where you want to compare each group with the others
Here you use the Tukey HSD test:
```{r}
tukey_comparisons <- df %>% tukey_hsd(activity ~ temperature)
tukey_comparisons
```

### For the situation where you want to compare one control group with multiple treatments
```{r}
anova.results <- dataset_treatment %>% anova_test(activity ~ treatment)
anova.results
```
NEED TO WRITE THIS


